Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-09-23 Thesis/Paper_III.tex
Final commit

\newcommand{\kpch}{\ensuremath{h^{-1} \, \text{kpc}}}
\newcommand{\kpc}{\ensuremath{\text{kpc}}}
\newcommand{\solarmh}{\ensuremath{h^{-1} \, M_\odot}}
\newcommand{\solarm}{M_\odot}
\chapter{Forming Vertical Structure in $\Lambda$CDM Discs}\label{ch:paper_iii}
\textbf{This chapter contains a draft of a paper in preparation to be submitted to Monthly Notices of the Royal Astronomical Society. The working title for the paper is ``Vertical Structure and Kicked-Up Stars in $\Lambda$CDM Galaxies". }

@@

We perform numerical experiments where stellar discs are inserted into
dark matter halos in a $\Lambda$CDM simulation after which the
fully-live disc-halo systems are evolved until the present epoch. {Prior it its insertion, a disc is treated as a growing, rigid potential whose position and orientation are determined from rigid-body dynamics. When the disc goes live, the disc-halo system is in approximate equilibrium.} The
discs develop structures such as bars, warps, and bending waves.  Not
surprisingly, we find that when we insert the disc at a redshift
$z=1$, it develops more structure and exhibits more clearly signs of
disequilibrium then when it is inserted later.  We identify
populations of stars kicked out of the disc to high latitudes that are
qualitatively similar what is observed in the Milky Way. {An in-depth attempt is made to disentangle the effects of bar buckling, tidal fields due to 
the generally triaxial halo and from distant subhalos, and the
passage of subhaloes through the inner disc. }While our
results are consistent with the hypothesis the the kicked-out-disc was
produced by a very massive satellite galaxy, we also find that they
can arise from the interaction of the disc with its halo. We conclude
that in the Milky Way a wide variety of cosmological environments are
able to explain disc structures like Monoceros, A13, and TriAnd.

@@

The single-satellite encounter is a rather idealized scenario;
disc-environment interactions in real galaxies will, of course, be
more complicated. In particular, discs may be continually perturbed by
any number of satellites or dark matter subhaloes whose orbits take
them into the disc region of the galaxy. Early steps toward a more
realistic model were taken by \citet{font_2001} and
\citet{gauthier_2006} who allowed for a system of subhalos. In
particular, \citet{gauthier_2006} simulated an M31-like disc in a dark
halo where roughly 10\% of the smooth halo mass was replace by dark
subhalos with a mass spectrum motivated by cosmological $\Lambda$CDM
simulations. They found that one or more of the subhalos triggered the
formation of a bar. Further analyses of this simulation and a similar
one by \citet{chequers_2018} found that the system of subhalos could
excite a plethora of vertical structure. However, these simulations
failed to create the streams of disc stars described above.

@@

Finally, we note that dark halos in $\Lambda$CDM cosmologies are
thought to be aspherical and will therefore exert torques on the disc
that can cause it to bend and warp. The effect is due to a
misalignment of the disc symmetry axis with any of the approximate
symmetry axes of the halo. For example, \citet{hu_2016} embed discs
in triaxial haloes. In their appendix, strong warps can be seen in
discs which are substantially misaligned from their
haloes. \citet{gomez_2017} found that vertical structure was
ubiquitous in their hydrodynamical cosmological simulations, some of
which did not have discs in a stable orientation to their host halo.

@@

For our pure dark matter simulation, we implement the zoom-in
technique of \citet{KatzQuasarZoom} and \citet{NavarroWhiteZoom}.  We
select Milky Way-like halos with no major mergers and use the results
of \cite{onorbe_etal_2014} to ensure minimal contamination of
low-resolution particles.  Our choice of cosmological parameters is
based on the results from Planck 2013 \citep{planck_2014}:
$H_0=67.9\,{\rm km\,s}^{-1}\,{\rm kpc}^{-1}$, $\Omega_b = 0.0481$,
$\Omega_0 = 0.306$, $\Omega_\Lambda = 0.694$, $\sigma_8 = 0.827$, and
$n_s = 0.962$. N-body initial conditions for the dark matter particles
are generated with the \textsc{music} code \citep{music}.

@@

We present a concise description of our cosmological models in Table
\ref{tab:sims_iii}. The haloes are chosen such that the total circular
speed curves are similar to the Milky Way's. Note that the bauer2018b
simulations use a halo from \cite{bauer2018b}. The halos used in the
A, B, and C simulations are previously unpublished.  In the
simulations denoted as X.Early, the disc is inserted at $a_l=0.5$; for
those denoted X.Lake, $a_l=0.75$.  One of the benefits of disc
insertion techniques is the ability to modify disc properties in the
same halo. We exploit this capability in our third suite, denoted as
X.Warm, where we increase the disc's dynamical temperature as
measured by the Toomre $Q$ parameter \citep{toomre_q}.

@@

\begin{sidewaystable}
\centering
\caption{Table of simulations run in the main experiment. $M_h$ is the
  virial mass of the host halo, $R_h$ is the NFW scale length of the
  host halo, $c$ is the NFW concentration parameter of the host halo,
  $a_l$ is the live disc insertion scale factor, $\delta
  \Theta_{h,20}$ is the halo misalignment angle for the 20 kpc shell,
  $N_d$ is the number of disc particles, $Q$ is the Toomre stability
  parameter of the disc at initialization, and $f_{kud}$ is the
  fraction of disc mass that gets classified as kicked-out-disk (KOD)
  stars measured at present day with $\eta=8$.} \label{tab:sims_iii}
\begin{tabular}{l l l l l l l l l l}
\hline
       & $M_h\,\,(\solarm)$ & $R_h\,\,(\kpc)$ & $c$ & $a_l$ &  $N_d$ & $Q$ & $\cos\delta \Theta_{h,20}$ & $f_{kud}$\\
\hline
bauer2018.Early   & $1.0 \times 10^{12}$ & 15.9 &  9.84 &   0.50 & $1.0 \times 10^6$ & 1.1 & 0.16 & $1.2 \times 10^{-2}$\\
bauer2018.Warm    & $1.0 \times 10^{12}$ & 15.9 &  9.84 &   0.50 & $1.0 \times 10^6$ & 1.1 & 0.16 & $1.1 \times 10^{-2}$\\
bauer2018.Late    & $1.2 \times 10^{12}$ & 15.2 &  15.4 &   0.75 & $1.0 \times 10^6$ & 1.8 & 0.93 &$4.2 \times 10^{-3}$\\
A.Early           & $9.4 \times 10^{11}$ & 17.8 &  8.56 &   0.50 & $3.5 \times 10^6$ & 1.3 & 0.97 &$3.7 \times 10^{-3}$\\
A.Warm            & $9.4 \times 10^{11}$ & 17.8 &  8.56 &   0.50 & $3.5 \times 10^6$ & 1.8 & 0.97 & $1.6 \times 10^{-3}$\\
A.Late            & $1.1 \times 10^{12}$ & 18.5 &  12.1 &   0.75 & $3.5 \times 10^6$ & 1.3 & 0.98 & $6.8 \times 10^{-5}$\\
B.Early           & $5.9 \times 10^{11}$ & 10.4 &  12.6 &   0.50 & $3.5 \times 10^6$ & 1.3 & 0.91 & $8.5 \times 10^{-4}$\\
B.Warm            & $5.9 \times 10^{11}$ & 10.4 &  12.6 &   0.50 & $3.5 \times 10^6$ & 1.8 & 0.91 & $1.2 \times 10^{-3}$\\
B.Late            & $1.1 \times 10^{12}$ & 33.2 &  6.70 &   0.75 & $3.5 \times 10^6$ & 1.3 & 0.99 & $9.2 \times 10^{-4}$\\
C.Early           & $4.3 \times 10^{11}$ & 8.33 &  14.1 &   0.50 & $3.5 \times 10^6$ & 1.3 & 0.84 & $5.3 \times 10^{-3}$\\
C.Warm            & $4.3 \times 10^{11}$ & 8.33 &  14.1 &   0.50 & $3.5 \times 10^6$ & 1.8 & 0.84 & $5.6 \times 10^{-3}$\\
C.Late            & $6.7 \times 10^{11}$ & 13.6 &  14.0 &   0.75 & $3.5 \times 10^6$ & 1.3 & 0.92 & $1.0 \times 10^{-4}$\\
\hline
\end{tabular}

@@

In Table \ref{tab:substructure}, we list the five most massive (at
initialization) substructures for each of our simulations while the
orbits of the two most massive substructures are shown in
Fig. \ref{fig:substructure_orbits}.  The orbits present a wide variety
of behaviour. For example, the subhaloes in bauer2018b.Early spend
most of their lives in the outer halo on roughly tangential orbits. By
contrast, the second most mass subhalo in C.late orbits to within
$\sim 20\,{\rm kpc}$ of the disc centre. The subhalos in A.early are
more massive but never reach to within $30\,{\rm kpc}$.

@@

We now turn our attention to the dynamics of KOD stars, identified via
Eq. \ref{eq:kud_def} with $\eta=8$ in the final
snapshot. Fig. \ref{fig:kud_stars_final_ics} shows the KOD stars in
our fiducial suite at the final snapshot. In addition, we show the
same stars traced back to their position in the initial snapshot. The
stars are color-coded by the time at which they are kicked out of the
disc.  We immediately see that the KOD structures vary markedly across
our simulations. For example, A.Early shows a relatively small KOD
population.  Moreover, the KOD structure it does have exists at very
low latitudes and was kicked out fairly early in the
simulation. B.Early has a few streams of KOD stars, which were kicked
out of the disc at roughly the time when the disc had its close
encounters with massive subhalos. Indeed, there appear to be two
distinct populations of KOD stars. Finally, bauer2018b.Early and
C.Early show the most significant populations of KOD stars.  These
simulations correspond to the simulations with the most significant
disc-halo misalignment and shortest $T_\tau$. In both cases, we see a
continuous range of times over which stars were kicked out. Moreover,
in C.Early, many of the KOD stars were originally in the inner disc.

@@

Perhaps the most novel effect that we have explored is the generation
of flapping modes from differential acceleration of the disc by
distant subhalos and other mass concentrations in a $\Lambda$CDM cosmology.
\citet{sellwood_1996} presented the first in-depth study
of this phenomenon in isolated galaxies. We find that the disc
flapping naturally arises in the cosmological environment and can be
due to one of three mechanisms: two-body interaction of the 
disc and inner halo with a subhalo, buckling of the bar,
and torquing of the disc by its smooth, triaxial halo.

@@

\section{Conclusions}\label{sec:conclusion}
{We have presented a suite of simulations in which stellar discs are
inserted into cosmological haloes. These halos are, in general,
triaxial and clumpy and drive the disc from its initial axisymmetric
and near equilibrium state to one of disequilibrium. The key findings are summarized as:
\begin{itemize}
\item All discs all form bars, even though buckling can regulate their strengths.
\item All discs exhibit some kind of bending and warping.
\item Disc flapping is a major manifestation of disequilibrium in the cosmological environment.
\item Subhaloes with a mass lower than that of the disc are unlikely to generate KOD populations.
\end{itemize}}
{It is not surprising that} the discs in our simulations bend and warp. In
addition, we find large populations of stars kicked out of the disc.
These departures from planarity signal a state of disequilibrium {likely present in the Milky Way}.  We
have identified several mechanisms for generating vertical structure, which
include the well-studied interaction of the disc with a massive
satellite, differential acceleration of the disc due to tidal fields
of the halo or a distant massive satellite, and the buckling bar.
The first two mechanisms can produce large warps (amplitudes of
order 1-3 kpc) in the outer disc along with smaller amplitude 
waves that extend further in. On the other hand, the buckling bar
is efficient at producing bending waves and corrugation patterns 
in the inner disc.

>>>>

\newcommand{\kpch}{\ensuremath{h^{-1} \, \text{kpc}}}
\newcommand{\kpc}{\ensuremath{\text{kpc}}}
\newcommand{\solarmh}{\ensuremath{h^{-1} \, M_\odot}}
\newcommand{\solarm}{M_\odot}
\chapter{Forming Vertical Structure in $\Lambda$CDM Discs}\label{ch:paper_iii}
\textbf{This chapter contains a draft of a paper submitted to Monthly Notices of the Royal Astronomical Society. Its current manuscript ID is MN-19-3241-MJ.}

@@

We perform numerical experiments where stellar discs are inserted into
dark matter haloes in a $\Lambda$CDM simulation after which the
fully-live disc-halo systems are evolved until the present epoch. {Prior it its insertion, a disc is treated as a growing, rigid potential whose position and orientation are determined from rigid-body dynamics. When the disc goes live, the disc-halo system is in approximate equilibrium.} The
discs develop structures such as bars, warps, and bending waves.  Not
surprisingly, we find that when we insert the disc at a redshift
$z=1$, it develops more structure and exhibits more clearly signs of
disequilibrium then when it is inserted later.  We identify
populations of stars kicked out of the disc to high latitudes that are
qualitatively similar what is observed in the Milky Way. {An in-depth attempt is made to disentangle the effects of bar buckling, tidal fields due to 
the generally triaxial halo and from distant subhaloes, and the
passage of subhaloes through the inner disc. }While our
results are consistent with the hypothesis the the kicked-out-disc was
produced by a very massive satellite galaxy, we also find that they
can arise from the interaction of the disc with its halo. We conclude
that in the Milky Way a wide variety of cosmological environments are
able to explain disc structures like Monoceros, A13, and TriAnd.

@@

The single-satellite encounter is a rather idealized scenario;
disc-environment interactions in real galaxies will, of course, be
more complicated. In particular, discs may be continually perturbed by
any number of satellites or dark matter subhaloes whose orbits take
them into the disc region of the galaxy. Early steps toward a more
realistic model were taken by \citet{font_2001} and
\citet{gauthier_2006} who allowed for a system of subhaloes. In
particular, \citet{gauthier_2006} simulated an M31-like disc in a dark
halo where roughly 10\% of the smooth halo mass was replace by dark
subhaloes with a mass spectrum motivated by cosmological $\Lambda$CDM
simulations. They found that one or more of the subhaloes triggered the
formation of a bar. Further analyses of this simulation and a similar
one by \citet{chequers_2018} found that the system of subhaloes could
excite a plethora of vertical structure. However, these simulations
failed to create the streams of disc stars described above.

@@

Finally, we note that dark haloes in $\Lambda$CDM cosmologies are
thought to be aspherical and will therefore exert torques on the disc
that can cause it to bend and warp. The effect is due to a
misalignment of the disc symmetry axis with any of the approximate
symmetry axes of the halo. For example, \citet{hu_2016} embed discs
in triaxial haloes. In their appendix, strong warps can be seen in
discs which are substantially misaligned from their
haloes. \citet{gomez_2017} found that vertical structure was
ubiquitous in their hydrodynamical cosmological simulations, some of
which did not have discs in a stable orientation to their host halo.

@@

For our pure dark matter simulation, we implement the zoom-in
technique of \citet{KatzQuasarZoom} and \citet{NavarroWhiteZoom}.  We
select Milky Way-like haloes with no major mergers and use the results
of \cite{onorbe_etal_2014} to ensure minimal contamination of
low-resolution particles.  Our choice of cosmological parameters is
based on the results from Planck 2013 \citep{planck_2014}:
$H_0=67.9\,{\rm km\,s}^{-1}\,{\rm kpc}^{-1}$, $\Omega_b = 0.0481$,
$\Omega_0 = 0.306$, $\Omega_\Lambda = 0.694$, $\sigma_8 = 0.827$, and
$n_s = 0.962$. N-body initial conditions for the dark matter particles
are generated with the \textsc{music} code \citep{music}.

@@

We present a concise description of our cosmological models in Table
\ref{tab:sims_iii}. The haloes are chosen such that the total circular
speed curves are similar to the Milky Way's. Note that the bauer2018b
simulations use a halo from \cite{bauer2018b}. The haloes used in the
A, B, and C simulations are previously unpublished.  In the
simulations denoted as X.Early, the disc is inserted at $a_l=0.5$; for
those denoted X.Lake, $a_l=0.75$.  One of the benefits of disc
insertion techniques is the ability to modify disc properties in the
same halo. We exploit this capability in our third suite, denoted as
X.Warm, where we increase the disc's dynamical temperature as
measured by the Toomre $Q$ parameter \citep{toomre_q}.

@@

\begin{sidewaystable}
\centering
\caption{Table of simulations run in the main experiment. $M_h$ is the
  virial mass of the host halo, $R_h$ is the NFW scale length of the
  host halo, $c$ is the NFW concentration parameter of the host halo,
  $a_l$ is the live disc insertion scale factor, $\delta
  \Theta_{h,20}$ is the halo misalignment angle for the 20 kpc shell,
  $N_d$ is the number of disc particles, $Q$ is the Toomre stability
  parameter of the disc at initialization, and $f_{kud}$ is the
  fraction of disc mass that gets classified as kicked-out-disc (KOD)
  stars measured at present day with $\eta=8$.} \label{tab:sims_iii}
\begin{tabular}{l l l l l l l l l l}
\hline
       & $M_h\,\,(\solarm)$ & $R_h\,\,(\kpc)$ & $c$ & $a_l$ &  $N_d$ & $Q$ & $\cos\delta \Theta_{h,20}$ & $f_{kud}$\\
\hline
bauer2018.Early   & $1.0 \times 10^{12}$ & 15.9 &  9.84 &   0.50 & $1.0 \times 10^6$ & 1.1 & 0.16 & $1.2 \times 10^{-2}$\\
bauer2018.Warm    & $1.0 \times 10^{12}$ & 15.9 &  9.84 &   0.50 & $1.0 \times 10^6$ & 1.1 & 0.16 & $1.1 \times 10^{-2}$\\
bauer2018.Late    & $1.2 \times 10^{12}$ & 15.2 &  15.4 &   0.75 & $1.0 \times 10^6$ & 1.8 & 0.93 &$4.2 \times 10^{-3}$\\
A.Early           & $9.4 \times 10^{11}$ & 17.8 &  8.56 &   0.50 & $3.5 \times 10^6$ & 1.3 & 0.97 &$3.7 \times 10^{-3}$\\
A.Warm            & $9.4 \times 10^{11}$ & 17.8 &  8.56 &   0.50 & $3.5 \times 10^6$ & 1.8 & 0.97 & $1.6 \times 10^{-3}$\\
A.Late            & $1.1 \times 10^{12}$ & 18.5 &  12.1 &   0.75 & $3.5 \times 10^6$ & 1.3 & 0.98 & $6.8 \times 10^{-5}$\\
B.Early           & $5.9 \times 10^{11}$ & 10.4 &  12.6 &   0.50 & $3.5 \times 10^6$ & 1.3 & 0.91 & $8.5 \times 10^{-4}$\\
B.Warm            & $5.9 \times 10^{11}$ & 10.4 &  12.6 &   0.50 & $3.5 \times 10^6$ & 1.8 & 0.91 & $1.2 \times 10^{-3}$\\
B.Late            & $1.1 \times 10^{12}$ & 33.2 &  6.70 &   0.75 & $3.5 \times 10^6$ & 1.3 & 0.99 & $9.2 \times 10^{-4}$\\
C.Early           & $4.3 \times 10^{11}$ & 8.33 &  14.1 &   0.50 & $3.5 \times 10^6$ & 1.3 & 0.84 & $5.3 \times 10^{-3}$\\
C.Warm            & $4.3 \times 10^{11}$ & 8.33 &  14.1 &   0.50 & $3.5 \times 10^6$ & 1.8 & 0.84 & $5.6 \times 10^{-3}$\\
C.Late            & $6.7 \times 10^{11}$ & 13.6 &  14.0 &   0.75 & $3.5 \times 10^6$ & 1.3 & 0.92 & $1.0 \times 10^{-4}$\\
\hline
\end{tabular}

@@

In Table \ref{tab:substructure}, we list the five most massive (at
initialization) substructures for each of our simulations while the
orbits of the two most massive substructures are shown in
Fig. \ref{fig:substructure_orbits}.  The orbits present a wide variety
of behaviour. For example, the subhaloes in bauer2018b.Early spend
most of their lives in the outer halo on roughly tangential orbits. By
contrast, the second most mass subhalo in C.late orbits to within
$\sim 20\,{\rm kpc}$ of the disc centre. The subhaloes in A.early are
more massive but never reach to within $30\,{\rm kpc}$.

@@

We now turn our attention to the dynamics of KOD stars, identified via
Eq. \ref{eq:kud_def} with $\eta=8$ in the final
snapshot. Fig. \ref{fig:kud_stars_final_ics} shows the KOD stars in
our fiducial suite at the final snapshot. In addition, we show the
same stars traced back to their position in the initial snapshot. The
stars are color-coded by the time at which they are kicked out of the
disc.  We immediately see that the KOD structures vary markedly across
our simulations. For example, A.Early shows a relatively small KOD
population.  Moreover, the KOD structure it does have exists at very
low latitudes and was kicked out fairly early in the
simulation. B.Early has a few streams of KOD stars, which were kicked
out of the disc at roughly the time when the disc had its close
encounters with massive subhaloes. Indeed, there appear to be two
distinct populations of KOD stars. Finally, bauer2018b.Early and
C.Early show the most significant populations of KOD stars.  These
simulations correspond to the simulations with the most significant
disc-halo misalignment and shortest $T_\tau$. In both cases, we see a
continuous range of times over which stars were kicked out. Moreover,
in C.Early, many of the KOD stars were originally in the inner disc.

@@

Perhaps the most novel effect that we have explored is the generation
of flapping modes from differential acceleration of the disc by
distant subhaloes and other mass concentrations in a $\Lambda$CDM cosmology.
\citet{sellwood_1996} presented the first in-depth study
of this phenomenon in isolated galaxies. We find that the disc
flapping naturally arises in the cosmological environment and can be
due to one of three mechanisms: two-body interaction of the 
disc and inner halo with a subhalo, buckling of the bar,
and torquing of the disc by its smooth, triaxial halo.

@@

\section{Conclusions}\label{sec:conclusion}
{We have presented a suite of simulations in which stellar discs are
inserted into cosmological haloes. These haloes are, in general,
triaxial and clumpy and drive the disc from its initial axisymmetric
and near equilibrium state to one of disequilibrium. The key findings are summarized as:
\begin{itemize}
\item All discs all form bars, even though buckling can regulate their strengths.
\item All discs exhibit some kind of bending and warping.
\item Disc flapping is a major manifestation of disequilibrium in the cosmological environment.
\item Subhaloes with a mass lower than that of the disc are unlikely to generate KOD populations.
\end{itemize}}
{It is not surprising that} the discs in our simulations bend and warp. In
addition, we find large populations of stars kicked out of the disc.
These departures from planarity signal a state of disequilibrium {likely present in the Milky Way}.  We
have identified several mechanisms for generating vertical structure, which
include the well-studied interaction of the disc with a massive
satellite, differential acceleration of the disc due to tidal fields
of the halo or a distant massive satellite, and the buckling bar.
The first two mechanisms can produce large warps (amplitudes of
order 1-3 kpc) in the outer disc along with smaller amplitude 
waves that extend further in. On the other hand, the buckling bar
is efficient at producing bending waves and corrugation patterns 
in the inner disc.

### 2019-09-23 Thesis/Thesis.tex
Final commit

\author{J S Bauer}

\dept{Department of Physics}
\degree{Doctor of Philosophy}

@@

Chapter~\ref{ch:paper_iii} contains a draft of a paper in preparation to be submitted to Monthly Notices of the Royal Astronomical Society. The working title for the paper is ``Vertical Structure and Kicked-Up Stars in $\Lambda$CDM Galaxies". For this work, I performed all of the simulations and writing. The analysis was handled primarily by me under the direction of Lawrence M. Widrow.

>>>>

\author{Jacob S. Bauer}

\dept{Department of Physics, Engineering Physics \& Astronomy}
\degree{Doctor of Philosophy}

@@

Chapter~\ref{ch:paper_iii} contains a draft of a paper submitted to Monthly Notices of the Royal Astronomical Society. Its current manuscript ID is MN-19-3241-MJ. For this work, I performed all of the simulations and writing. The analysis was handled primarily by me under the direction of Lawrence M. Widrow.
